### 2025-02-05 paper/body_ecml.tex
vamos

Let \(M_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a supervised
classification model that maps from intuts to representations
\(\phi(x;theta)\) and finally to outputs. Assume that for any given
input-output pair \(\{\mathbf{x},\mathbf{y}\}\) there exists a
counterfactual
\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: M_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = M_\theta(\mathbf{x})\)
where \(\mathbf{y}^{+}\) denotes some target output. We say that
\(M_\theta\) is \textbf{explainable} to the extent that faithfully
generated counterfactuals are:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plausible: \(\mathbf{x}^{\prime} \sim_{p} \mathcal{X}|\mathbf{y}^{+}\)
  with \(p \rightarrow 1\) as defined in \cite{altmeyer2024faithful}.
\item
  Actionable: permutations \(\Delta\) are subject to actionability
  constraints.
\end{enumerate}

@@

Actionability constraints in Definition~\ref{def-explainability} vary
and depend on the context in which \(M_\theta\) is deployed. In this
work, we focus on domain and mutability constraints for individual
features. We also limit ourselves to classification tasks in this work,
a limitation that we discuss in Section~\ref{sec-discussion}.

@@

\begin{equation}\phantomsection\label{eq-obj}{
\text{yloss}(M_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(M_\theta({x^\prime}),y)
}\end{equation}

where \(\text{yloss}(\cdot)\) denotes any conventional classification
loss function (e.g.~crossentropy) that induces discriminative
(predictive) performance. The two additional components in
Equation~\ref{eq-obj} are explained in more detail below. For now they
can be sufficiently described as inducing explainability directly and
indirectly by penalizing: 1) the contrastive divergence,
\(\text{div}(\cdot)\), between counterfactuals \(x^\prime\) and observed
samples \(x\) and, 2) the adversarial loss, \(\text{advloss}(.)\), with
respect to counterfactuals. The tradeoff between the different
components can be governed by adjusting the strengths of the penalties
\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).

@@

@@

A reasonable choice for the latter is to define
\(\text{advloss}(M_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(M_\theta({x^\prime}),y)\)

>>>>

Let \(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a
supervised classification model that maps from the \(D\)-dimensional
input space \(\mathcal{X}\) to representations
\(\phi(\mathbf{x};\theta)\) and finally to the \(K\)-dimensional output
space \(\mathcal{Y}\). Assume that for any given input-output pair
\(\{\mathbf{x},\mathbf{y}\}_i\) there exists a counterfactual
\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})\)
where \(\mathbf{y}^{+}\) denotes some target output. We say that
\(\mathbf{M}_\theta\) is \textbf{explainable} to the extent that
faithfully generated counterfactuals are plausible (i.e.~consistent with
the data) and actionable. Formally, we define these properties as
follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  (Plausibility)
  \(\int^{A} p(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
  where \(A\) is some small region around \(\mathbf{x}^{\prime}\).
\item
  (Actionability) Permutations \(\Delta\) are subject to actionability
  constraints.
\end{enumerate}

We consider counterfactuals as faithful to the extent that they are
consistent with what the model has learned about the input data. Let
\(p_\theta(\mathbf{x}|\mathbf{y}^{+})\) denote the conditional posterior
over inputs, then formally:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  (Faithfulness)
  \(\int^{A} p_\theta(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
  where \(A\) is defined as above.
\end{enumerate}

@@

The definitions of faithfulness and plausibility in
Definition~\ref{def-explainability} are the same as in
\cite{altmeyer2024faithful}, with adapted notation. Actionability
constraints in Definition~\ref{def-explainability} vary and depend on
the context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
focus on domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
for reasons discussed in Section~\ref{sec-discussion}.

@@

\begin{equation}\phantomsection\label{eq-obj}{
\text{yloss}(\mathbf{M}_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta({x^\prime}),y)
}\end{equation}

where \(\text{yloss}(\cdot)\) denotes any conventional classification
loss function (e.g.~cross-entropy) that induces discriminative
(predictive) performance. The two additional components in
Equation~\ref{eq-obj} are explained in more detail below. For now, they
can be sufficiently described as inducing explainability directly and
indirectly by penalizing: 1) the contrastive divergence,
\(\text{div}(\cdot)\), between counterfactuals \(x^\prime\) and observed
samples \(x\) and, 2) the adversarial loss, \(\text{advloss}(.)\), with
respect to counterfactuals. The tradeoff between the different
components can be governed by adjusting the strengths of the penalties
\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).

@@

\cite{grathwohl2020your} observe that any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on inputs and generate inputs.
They show that JEMs can be trained to perform well at both tasks by
directly maximizing the joint log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first factor can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
\cite{grathwohl2020your} minimize the contrastive divergence between
samples drawn from \(p_\theta(\mathbf{x})\) and training observations,
i.e.~samples from \(p(\mathbf{x})\).

A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
tend to do well with respect to the plausibility objective in
Definition~\ref{def-explainability}. If we consider samples drawn from
\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
finding, because the JEM objective effectively minimizes the divergence
between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
To generate samples, \cite{grathwohl2020your} rely on Stochastic
Gradient Langevin Dynamics (SGLD) using an uninformative prior for
initialization. This is where we depart from their methodology: instead
of generating samples through SGLD, we propose using counterfactual
explainers to generate counterfactuals for observed training samples.

This departure allows us to tap into the vast repertoire of explainers
that have been proposed in the literature to meet different desiderata.
Typically, these methods facilitate the imposition of domain and
mutability constraints, for example. In principle, any existing approach
for generating counterfactual explanations is viable, so long as it does
not violate the faithfulness condition.

@@

A reasonable choice for the latter is to define
\(\text{advloss}(\mathbf{M}_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(\mathbf{M}_\theta({x^\prime}),y)\)

### 2025-02-05 paper/preprint/paper.tex
vamos

Let \(M_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a supervised
classification model that maps from intuts to representations
\(\phi(x;theta)\) and finally to outputs. Assume that for any given
input-output pair \(\{\mathbf{x},\mathbf{y}\}\) there exists a
counterfactual
\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: M_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = M_\theta(\mathbf{x})\)
where \(\mathbf{y}^{+}\) denotes some target output. We say that
\(M_\theta\) is \textbf{explainable} to the extent that faithfully
generated counterfactuals are:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plausible: \(\mathbf{x}^{\prime} \sim_{p} \mathcal{X}|\mathbf{y}^{+}\)
  with \(p \rightarrow 1\) as defined in Altmeyer et al.
  (\citeproc{ref-altmeyer2024faithful}{2024}).
\item
  Actionable: permutations \(\Delta\) are subject to actionability
  constraints.
\end{enumerate}

@@

Actionability constraints in Definition~\ref{def-explainability} vary
and depend on the context in which \(M_\theta\) is deployed. In this
work, we focus on domain and mutability constraints for individual
features. We also limit ourselves to classification tasks in this work,
a limitation that we discuss in Section~\ref{sec-discussion}.

@@

\begin{equation}\phantomsection\label{eq-obj}{
\text{yloss}(M_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(M_\theta({x^\prime}),y)
}\end{equation}

where \(\text{yloss}(\cdot)\) denotes any conventional classification
loss function (e.g.~cross-entropy) that induces discriminative
(predictive) performance. The two additional components in
Equation~\ref{eq-obj} are explained in more detail below. For now they
can be sufficiently described as inducing explainability directly and
indirectly by penalizing: 1) the contrastive divergence,
\(\text{div}(\cdot)\), between counterfactuals \(x^\prime\) and observed
samples \(x\) and, 2) the adversarial loss, \(\text{advloss}(.)\), with
respect to counterfactuals. The tradeoff between the different
components can be governed by adjusting the strengths of the penalties
\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).

@@

Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) observe that
any classifier can be re-interpreted as a joint energy-based model (JEM)
that learns to discriminate output classes conditional on inputs and
generate inputs. They show that JEMs can perform well at both tasks by
directly optimizing the joint distribution
\(\log p_\theta(x,y)=\log p_\theta(y|x) + \log p_\theta(x)\). The first
factor The distribution over inputs is learned via contrastive
divergence:

Considered through the lens of explainability, JEMs learn plausible
explanations.

>>>>

Let \(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a
supervised classification model that maps from the \(D\)-dimensional
input space \(\mathcal{X}\) to representations
\(\phi(\mathbf{x};\theta)\) and finally to the \(K\)-dimensional output
space \(\mathcal{Y}\). Assume that for any given input-output pair
\(\{\mathbf{x},\mathbf{y}\}_i\) there exists a counterfactual
\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})\)
where \(\mathbf{y}^{+}\) denotes some target output. We say that
\(\mathbf{M}_\theta\) is \textbf{explainable} to the extent that
faithfully generated counterfactuals are plausible (i.e.~consistent with
the data) and actionable. Formally, we define these properties as
follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  (Plausibility)
  \(\int^{A} p(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
  where \(A\) is some small region around \(\mathbf{x}^{\prime}\).
\item
  (Actionability) Permutations \(\Delta\) are subject to actionability
  constraints.
\end{enumerate}

We consider counterfactuals as faithful to the extent that they are
consistent with what the model has learned about the input data. Let
\(p_\theta(\mathbf{x}|\mathbf{y}^{+})\) denote the conditional posterior
over inputs, then formally:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  (Faithfulness)
  \(\int^{A} p_\theta(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
  where \(A\) is defined as above.
\end{enumerate}

@@

The definitions of faithfulness and plausibility in
Definition~\ref{def-explainability} are the same as in Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}), with adapted notation.
Actionability constraints in Definition~\ref{def-explainability} vary
and depend on the context in which \(\mathbf{M}_\theta\) is deployed. In
this work, we focus on domain and mutability constraints for individual
features \(x_d\) for \(d=1,...,D\). We limit ourselves to classification
tasks for reasons discussed in Section~\ref{sec-discussion}.

@@

\begin{equation}\phantomsection\label{eq-obj}{
\text{yloss}(\mathbf{M}_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta({x^\prime}),y)
}\end{equation}

where \(\text{yloss}(\cdot)\) denotes any conventional classification
loss function (e.g.~cross-entropy) that induces discriminative
(predictive) performance. The two additional components in
Equation~\ref{eq-obj} are explained in more detail below. For now, they
can be sufficiently described as inducing explainability directly and
indirectly by penalizing: 1) the contrastive divergence,
\(\text{div}(\cdot)\), between counterfactuals \(x^\prime\) and observed
samples \(x\) and, 2) the adversarial loss, \(\text{advloss}(.)\), with
respect to counterfactuals. The tradeoff between the different
components can be governed by adjusting the strengths of the penalties
\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).

@@

Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) observe that
any classifier can be re-interpreted as a joint energy-based model (JEM)
that learns to discriminate output classes conditional on inputs and
generate inputs. They show that JEMs can be trained to perform well at
both tasks by directly maximizing the joint log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first factor can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimize the
contrastive divergence between samples drawn from
\(p_\theta(\mathbf{x})\) and training observations, i.e.~samples from
\(p(\mathbf{x})\).

A key empirical finding in Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}) was that JEMs tend to do
well with respect to the plausibility objective in
Definition~\ref{def-explainability}. If we consider samples drawn from
\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
finding, because the JEM objective effectively minimizes the divergence
between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
To generate samples, Grathwohl et al.
(\citeproc{ref-grathwohl2020your}{2020}) rely on Stochastic Gradient
Langevin Dynamics (SGLD) using an uninformative prior for
initialization. This is where we depart from their methodology: instead
of generating samples through SGLD, we propose using counterfactual
explainers to generate counterfactuals for observed training samples.

This departure allows us to tap into the vast repertoire of explainers
that have been proposed in the literature to meet different desiderata.
Typically, these methods facilitate the imposition of domain and
mutability constraints, for example. In principle, any existing approach
for generating counterfactual explanations is viable, so long as it does
not violate the faithfulness condition.

### 2025-02-05 paper/sections/method.qmd
vamos

Let $M_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from intuts to representations $\phi(x;theta)$ and finally to outputs. Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: M_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = M_\theta(\mathbf{x})$ where $\mathbf{y}^{+}$ denotes some target output. We say that $M_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are:

1. Plausible: $\mathbf{x}^{\prime} \sim_{p} \mathcal{X}|\mathbf{y}^{+}$ with $p \rightarrow 1$  as defined in @altmeyer2024faithful.
2. Actionable: permutations $\Delta$ are subject to actionability constraints.

@@

Actionability constraints in @def-explainability vary and depend on the context in which $M_\theta$ is deployed. In this work, we focus on domain and mutability constraints for individual features. We also limit ourselves to classification tasks in this work, a limitation that we discuss in @sec-discussion.

@@

$$
\text{yloss}(M_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(M_\theta({x^\prime}),y)
$$ {#eq-obj}

where $\text{yloss}(\cdot)$ denotes any conventional classification loss function (e.g. cross-entropy) that induces discriminative (predictive) performance. The two additional components in @eq-obj are explained in more detail below. For now they can be sufficiently described as inducing explainability directly and indirectly by penalizing: 1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $x^\prime$ and observed samples $x$ and, 2) the adversarial loss, $\text{advloss}(.)$, with respect to counterfactuals. The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.

@@

@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on inputs and generate inputs. They show that JEMs can perform well at both tasks by directly optimizing the joint distribution $\log p_\theta(x,y)=\log p_\theta(y|x) + \log p_\theta(x)$. The first factor 
The distribution over inputs is learned via contrastive divergence:

Considered through the lens of explainability, JEMs learn plausible explanations.

@@

A reasonable choice for the latter is to define $\text{advloss}(M_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(M_\theta({x^\prime}),y)$

>>>>

Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$. Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\mathbf{y}^{+}$ denotes some target output. We say that $\mathbf{M}_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are plausible (i.e. consistent with the data) and actionable. Formally, we define these properties as follows:

1. (Plausibility) $\int^{A} p(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
2. (Actionability) Permutations $\Delta$ are subject to actionability constraints.

We consider counterfactuals as faithful to the extent that they are consistent with what the model has learned about the input data. Let $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denote the conditional posterior over inputs, then formally:

3. (Faithfulness) $\int^{A} p_\theta(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.

@@

The definitions of faithfulness and plausibility in @def-explainability are the same as in @altmeyer2024faithful, with adapted notation. Actionability constraints in @def-explainability vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we focus on domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We limit ourselves to classification tasks for reasons discussed in @sec-discussion.

@@

$$
\text{yloss}(\mathbf{M}_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta({x^\prime}),y)
$$ {#eq-obj}

where $\text{yloss}(\cdot)$ denotes any conventional classification loss function (e.g. cross-entropy) that induces discriminative (predictive) performance. The two additional components in @eq-obj are explained in more detail below. For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: 1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $x^\prime$ and observed samples $x$ and, 2) the adversarial loss, $\text{advloss}(.)$, with respect to counterfactuals. The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.

@@

@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on inputs and generate inputs. They show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$. The first factor can be optimized using conventional cross-entropy as in @eq-obj. To optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between samples drawn from $p_\theta(\mathbf{x})$ and training observations, i.e. samples from $p(\mathbf{x})$.

A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in @def-explainability. If we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals, this is an expected finding, because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization. This is where we depart from their methodology: instead of generating samples through SGLD, we propose using counterfactual explainers to generate counterfactuals for observed training samples.

This departure allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata. Typically, these methods facilitate the imposition of domain and mutability constraints, for example. In principle, any existing approach for generating counterfactual explanations is viable, so long as it does not violate the faithfulness condition.

@@

A reasonable choice for the latter is to define $\text{advloss}(\mathbf{M}_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(\mathbf{M}_\theta({x^\prime}),y)$
